Reproducible Research: Peer-graded Assesment 1
==============================================

## Loading Data and exploratory analysis of the data
The data is first read and summarised
```{r opts_chunks$set(echo=TRUE)}
activity <- read.csv("activity.csv")
activity$date <- as.Date(as.character(activity$date))
head(activity)
summary(activity)
pairs(activity)
```

## 1.What is mean total number of steps taken per day?
The missing values in the data set are removed and the data is aggregated on steps and day using sum. A histogram is plotted on steps per day. 
```{r}
stepsperday<- aggregate(steps~date, activity,sum, na.rm=T)
hist(stepsperday$steps, xlab="No. of steps taken per day", main="Histogram of no. of steps taken per day")
```

The mean of total no. of steps taken is given by
```{r}
meansteps <- mean(stepsperday$steps)
meansteps
```
The median no. of steps taken is given by:
```{r}
mediansteps <- median(stepsperday$steps) 
```

## 2.What is the average daily activity pattern?
To find the average daily activity pattern, the data is aggregated on steps and interval using mean and missing values are removed. A line graph is drawn with interval on X-axis and steps taken on Y- axis.
```{r}
act2 <- aggregate(steps~interval, activity, mean, na.rm=T)
plot(steps~interval, act2, type="l", xlab="Interval", ylab="Steps taken", main="Line graph of average steps taken per interval")
```

To find the interval where maximum steps were taken:
```{r}
maxinterval <- act2[which.max(act2$steps),]$interval
maxinterval
```

## 3.Imputing missing values
a) Calculate and report total number of missing values
```{r}
meansteps <- mean(stepsperday$steps)
meansteps
```

b)Devise a strategy for filling in all of the missing values in the dataset. Here we use the mean of the interval to fill the missing  values. A function to fill the missing values is written below.
```{r}
impute_val <- function(interval) {
  act2[act2$interval==interval,]$steps
}
```
c)The data is then copied into another variable and the copy of the the data is run through the function to fill the missing values. 
```{r}
act_noNA <- activity
for(i in 1:nrow(act_noNA)) {
  if(is.na(act_noNA[i,]$steps)) {
    act_noNA[i,]$steps <- impute_val(act_noNA[i,]$interval)
  }
}
```
d) A histogram of the no. of steps taken per day is made.
```{r}
totalsteps_noNA <- aggregate(steps~date, act_noNA, sum)
hist(totalsteps_noNA$steps, xlab="No. of steps taken", main="Histogram of no. of  steps taken")
``` 

Calculating the mean no. of steps taken
```{r}
meansteps_noNA <- mean(totalsteps_noNA$steps)
meansteps_noNA
```
The mean **does not change** even after imputing the missing values.
  
Calculating the median no. of steps taken.
```{r}
mediansteps_noNA <- median(totalsteps_noNA$steps)
mediansteps_noNA
```
The median changes by about **0.1%** after imputing the missing values.

## Are there differences in activity patterns between weekdays and weekends?
a)Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
act_noNA$date <- as.Date(strptime(act_noNA$date,forma="%Y-%m-%d"))
act_noNA$day <- weekdays(act_noNA$date)
for(i in 1:nrow(act_noNA)) {
    if(act_noNA[i,]$day=="Saturday"|act_noNA[i,]$day=="Sunday") {
      act_noNA[i,]$day <- "Weekend"
    }
    else {
      act_noNA[i,]$day <- "Weekday"
    }
}
stepsbyday <- aggregate(act_noNA$steps ~ act_noNA$interval + act_noNA$day, act_noNA, mean)
```
b)Make a panel plot containing a time series plot (i.e. type = "l"\color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
names(stepsbyday) <- c("interval", "day", "steps")
library(lattice)
xyplot(steps~interval|day, stepsbyday, layout=c(1,2),type="l",xlab="Interval", ylab="No. of steps")
```

